Determining Bravais Lattice Types

### 2025-04-02 SELLA/Sella.tex
udate and add References for Web Site
Commandusage.html -- remove mentinos of PREFIX.
LinearAxis, Claude's attempt to help me write an article. SELLA -- updating the SELLA article. More to do

\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\scalarsub}[2]{$#1_#2$}
\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}

@@

	{\LARGE \emph{\today}} \\
\title{SELLA - A Program for Determining Bravais Lattice Types}
%\shorttitle{SELLA  Determining Bravais Lattice Types}

@@

\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
\author[b]{Herbert J.}{Bernstein}
\author[c]{Nicholas K.}{Sauter}

@@

\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}, Rochester Institute of Technology, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
\aff[c]{Lawrence Berkeley National Laboratory, 1 Cyclotron Rd., Berkeley, CA, 94720 \country{USA}}

@@

\begin{abstract}
We introduce a new Bravais lattice determination algorithm. SELLA is a 
straight-forward algorithm and a program for determining Bravais lattice type 
based on Selling (Delone) reduction. It is a complete, closed solution, 
and it provides a clear metric of fit to each type.
\end{abstract}

{\bf Note:}  Boris Delaunay in his later publications used the Russian version 
of his surname: Delone. We will follow that choice.\\

@@

We introduce a new Bravais lattice determination algorithm. The Bravais 
lattice types were created by \citeasnoun{Bravais1850}. \citeasnoun{Delaunay1932} 
and \citeasnoun{Niggli1928} developed methods for the identification 
of the Bravais lattice type of a crystal using the measured unit cell 
dimensions; their methods were exact only if the cell parameters 
exactly corresponded to the actual type \cite{Patterson1957}. 
Delone discussed the issues of scalars that have nearly zero values, 
but the broader issues of other measurement errors were not 
discussed. \citeasnoun{andrews2014} review the literature on 
the efforts to create methods to utilize data that contain unavoidable 
measurement error.

SELLA is a algorithm and program for determining Bravais lattice 
type based on Selling (Delone) reduction. It is a complete, closed solution, and it provides a clear metric of fit to each type.

@@

\citeasnoun{Delaunay1932} found that the 14 Bravais lattice types 
could be allocated among 24 types that are determined by the 
type of Dirichlet unit cell (termed the Voronoi region \cite{Voronoi1908} or 
Dirichlet region \cite{dirichlet1850} by 
Delone). Delone used the convention of \citeasnoun{Bravais1850}
with the sides of the tetrahedron labeled by the labels of the six
Selling scalars. Figure \ref{fig:PQRSTU} shows the labeling used in
the International Tables for Crystallography \cite{Henry1952}; 
Table \ref{table:ScalarConcordance} shows other choices that have been used. The
 contents for each type are described in Figure \ref{fig:terminology}.

CHARACTER

\citeasnoun{Burzlaff1985} revised the labeling of the types.
Here we have returned to the numbering of the types to that 
of \citeasnoun{Delaunay1932}.
In some cases, Delone included two forms within a single 
type. \citeasnoun{Burzlaff1985}
renumbered by ordinals. Here we have labeled the Delone 
types that have multiple items as "A" and "B". Table \ref{table:DeloneTypeConcordance}
shows the various labelings. Figure \ref{fig:DeloneTypes} 
recapitulates the figure of \citeasnoun{Delaunay1932} using 
more modern notation. Table \ref{table:ScalarConcordance} 
lists the various labelings that have been 
used for the sides of the Bravais tetrahedron. 
Figure \ref{fig:basictriangle} shows the arrangement of the 
symbols as used in the International Tables for Crystallography.

@@

	\rotatebox{80}{\citeasnoun{Delaunay1932}}
		  & \rotatebox{80}{\citeasnoun{Henry1952}  }
		   & \rotatebox{80}{\citeasnoun{Patterson1957} }
		 & \rotatebox{80}{\citeasnoun{Burzlaff1985}}
		   &\rotatebox{80}{\citeasnoun{andrews2019} } \\	
			\midrule
		g&p&P&h$_{23}$&\scalarsub{s}{1}\\		
		k&r&R&h$_{12}$&\scalarsub{s}{3}\\		
		l&s&S&h$_{14}$&\scalarsub{s}{4}\\		
		m&t&T&h$_{24}$&\scalarsub{s}{5}\\		
		h&q&Q&h$_{13}$&\scalarsub{s}{2}\\		
		n&u&U&h$_{34}$&\scalarsub{s}{6}\\		
		\bottomrule
	\end{tabular}

@@

\begin{table}
	\begin{tabular}{ccccccccc}
		\toprule
		 \cite{Delaunay1932} & \cite{Burzlaff1985}   &this  \\
		&  & paper \\
		\midrule
		K$_I$ & K1 & C1 \\		
		K$_{III}$ & K2 & C3 \\		
		K$_V$ & K3 & C5 \\		
		Q$_I$ & Q1 & T1 \\		
		Q$_{II}$ & Q2 & T2 \\		
		Q$_V$ & Q3 & T5\\		
		R$_I$ & R1 & R1 \\		
		R$_{III}$ & R2 & R3 \\		
		O$_I$ & O1 & O1A  \\		
		O$_I$ & O2 & O1B  \\		
		O$_{II}$ & O3 & O2   \\		
		O$_{III}$ & O4 & O3   \\		
		O$_{IV}$ & O5 & O4   \\		
		O$_V$ & O6 & O5   \\		
		M$_I$ & M1 & M1A  \\		
		M$_I$ & M2 & M1B  \\		
		M$_{II_2}$  & M3 & M2A  \\		
		M$_{II}$ & M4 & M2B  \\		
		M$_{III}$ & M5 & M3  \\		
		M$_{IV}$ & M6 & M4  \\		
		T$_I$ & T1 & A1  \\ 		
		T$_{II}$ & T2 & A2 \\		
		T$_{III}$ & T3 & A3 \\		
		H$_{IV}$  & H1 & H4 \\ 
		\bottomrule
	\end{tabular}
~\\ ~\\
	\caption{The 24 Delone types. \cite{Delaunay1932}}
	\label{table:DeloneTypeConcordance}
\end{table}

\subsection{The space \SVI{}}
\cite{andrews2019} recast the Selling parameters to define a metric space, \SVI{}.
We use the base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell to define $\bf{d}=-(\bf{a}+\bf{b}+\bf{c})$.
A point ${s}$ in \SVI{} is defined as:\\
${s}$= [\scalarsub{s}{1}, \scalarsub{s}{2}, \scalarsub{s}{3},\scalarsub{s}{4}, \scalarsub{s}{5}, \scalarsub{s}{6}] \\
where \scalarsub{s}{1}=\vdotv{b}{c}, ~\scalarsub{s}{2}=\vdotv{a}{c}, ~\scalarsub{s}{3}=\vdotv{a}{b}, 
~\scalarsub{s}{4}=\vdotv{a}{d}, ~\scalarsub{s}{5}= \vdotv{b}{d}, and \scalarsub{s}{6}=\vdotv{c}{d}.

\subsection{Terminology of the Bravais lattice types}
The Bravais tetrahedron symbol. Figure \ref{fig:PQRSTU}  and Table \ref{table:DeloneTypeConcordance}\\
The different ways that the scalars have been labeled  Figure \ref{fig:terminology} \\
The table of the Delone types. Figure \ref{fig:DeloneTypes} \\

\section{Algorithms for creation of an exhaustive list of the polytopes of the Bravais types}

@@

We will use Selling reduction; Niggli reduction has a fundamental 
unit that is non-convex, and some boundaries of the fundamental 
unit are partly closed and partly open \cite{andrews2014} leading 
to considerable complexity. Because Niggli reduction divides the 
fundamental unit into separate all obtuse and all acute regions, 
 the transitions between those are quite complex \cite{andrews2014}. 
In space $\bf{S^{6}}$ \cite{andrews2019b}, the fundamental 
unit is the simple all-negative orthant of a 6-dimensional 
Cartesian space. Only two kinds of operations in the space 
will need to be considered: the 6 boundary transform operations 
at the boundaries of the orthant (zeros of an axis), and 
24 reflections \cite{andrews2019}

Twenty-four lattice types were enumerated by \citeasnoun{Delaunay1932}. 
Three of those are triclinic, and we will ignore them because all crystals 
can be indexed as triclinic. We 
shall need all representations of all non-triclinic lattice 
types within the $\bf{S^{6}}$ fundamental unit (all Selling-reduced 
cells) or those generated beyond the boundaries by boundary transforms. 
Three forms (M3, M2B, and O3) are reduced dimension boundaries that do
not correspond to unique Bravais types; any lattice that falls in those
three is indexed by one of the other types.

 Seven of the 21 Delone types do not have a zero in their $\bf{S^{6}}$ 
 representation; the presence of zero indicates that a form is 
 on the boundary of the $\bf{S^{6}}$ fundamental unit. Since a 
 point on the boundary transforms to another boundary 
 point \cite{andrews2019b}, we can generate the Virtual 
 Cartesian Point (VCP) \cite{andrews2019b} appropriate to 
 each zero in the $\bf{S^{6}}$ vector. Because the reduction
  operations do not commute, in the cases where there is more 
  than one zero in the vector, all orderings of applying 
  reductions to all of the zeros must be generated. (For example, 
  for two zeros, there will be six vectors produced.) Again, 
  there may be many duplicate results.

 Finally, the projectors must be computed for each of the many 
 sample vectors of the 21 types. We will also need the 
 ``perps'', the operators that give the normal vector to each 
 such manifold. For a particular polytope, the projector applied to the probe gives
 the least-squares best fit point within that polytope. Perps are computed by subtracting the 
 corresponding projector from the unit matrix. The norm of a perp times
 a probe is the distance from that polytope.

 \subsubsection{Generating sample vectors from all Bravais types}

 Step 1: Generate any random vector that contains no zero 
 values and no duplicate values.  For a later operation, 
 it should only contain values large enough that adding or
 subtracting the smallest
 representable floating point number will not change the 
 value in computer arithmetic ($DBL\_MIN$ in the C language).

 Step 2; For each Delone type, multiply its projector by 
 the vector from Step 1. Store  the results in a list 
 with entries for each type (see Table \ref{table:DeloneTypes} 
 for the projectors).

@@

a) If the vector contains no zeros, generate the 24 reflections
 and store in the list by type.

\textit{\textbf{LCA--need to normalize the vectors, just like in lattice matching. ? Can we use the 
lattice matching code here?}}

b) If the vector contains one zero, apply the boundary transform operation for that boundary
to the vector, and
 store the products of the vector and its 24 reflections in the list. Note, 
 this is where the choice of $DBL\_MIN$ has effect; adding or
 subtracting such a small value will not change
 any value but zero.

 c) If the vector contains more than one zero and \underline{not} $DBL\_MIN$, 
 first set the zero to $DBL\_MIN$,  and then return it to 
 step b). Then set the second zero to $DBL\_MIN$ (with the
 first zero still zero), and send that result to step b). 
 If there is a third zero,  process it the same way as two zeros and return to c).

@@

 \subsubsection{Generate projectors from sample vectors}

 Each vector in the list from the previous steps is a representative of one
 of the Delone types. Although a vector is a one-dimensional polytope,
 these vectors encode the information about the full polytopes of the
 Delone types in their zeros and their duplicate values.

@@

 The following steps will be repeated for each of the vectors in the
 list generated above. We generate a matrix \textit{m}.

 Step 1: Zero any elements that have been set to $DBL\_MIN$.

@@

\subsection {Measuring the fit}
The distance from the given polytope is just
the norm of the perp for that polytope applied to the probe.

@@

Three of the 21 non-triclinic Delone types are only boundaries of other types. In Figure 
\ref{fig:degenerate}, Delone types M2B and M3 are in the row for orthorhombic, and
O3 is in the row for rhombohedral and tetragonal. These types do not need to
be searched for in general surveys.

@@

Most of the proposals for Bravais lattice determination in the 
presence of experimental error do not have a clear proof that 
they are complete. For many years there were verbal complaints 
that available programs would not infrequently miss experimental 
cases (for example, TRACER \cite{lawton1965} and many anecdotal 
communications).  To our knowledge, only the method of 
\cite{tomiyasu2012} has clearly demonstrated completeness. 
The justification that Sella is complete has three parts.

First: We require that the cell being tested be reduced, and
all Bravais lattice types within the \SVI{} fundamental unit (all
\SVI{} scalars negative) be generated. Therefore,
 the 24 possible forms of a 
reduced cell will be close to one of the 24 copies of the types. 
For the reflections, we do not need to consider the case of 
points near a boundary that might be outside; the boundary 
transforms must be treated for that.

Second: In the case of the boundary transform operations, we take the case of a point that is infinitesimally distant from the polytope of a 
Bravais lattice type that has one zero in its scalars. Take 
the point in the polytope that is closest to the experimental 
point. We transform both by the corresponding boundary transform 
operation. We have generated a new pair of points in a 
different location, one of which is in some representation of 
the same Bravais lattice type. The points are still 
infinitesimally distant from each other.

Third: Assume we have a point that is outside the fundamental 
unit and is near an undiscovered polytope for a Bravais lattice 
type. If the point is near a boundary defined by a single 
zero, then we apply a boundary transform to the point and to 
the Bravais polytope, bringing them both into the fundamental 
unit. If all of the variations of all of the Bravais 
lattice types within and on the fundamental unit have 
been generated, then this putative undiscovered polytope 
has already been included in the list of possibilities. 
Similar arguments apply to the cases of two or three zeros.

If we apply all of the boundary transform operations to the 
representations of the Bravais lattice types, then 
there are no places that a point in the fundamental 
unit will fail to find the closest Bravais lattice 
(or lattices). As stated above, there are 239 
non-triclinic possible representations.

What might occur if some reduced point has an unreduced 
copy near the all-negative orthant?  Would it be 
possible for it to be close to one of the copies of one 
of the 21 Delone types? The reduced point is obviously in 
the all-negative orthant. Further, because all possible 
copies of all of the Delone type manifolds have been 
generated, the reduced point will be near a copy of the 
manifold it was near when it was unreduced.

@@

89539 unit cells were extracted from the Protein Data Bank 
(\citeasnoun{bernstein1977} and \citeasnoun{berman2000}). 
When evaluated with Sella, all but 57 were found at zero 
distance from the described crystal class. The 57 were 
identified as incorrect unit cells for the assigned 
crystal class; however, the distances from the assigned 
classes were small because the deviations from the 
appropriate parameters were modest (see Table \ref{tab:PDBCells}).
\begin{table}
	\label{tab:PDBCells}

@@

	\caption{}
\end{table}
\subsection{Testing for continuity}

The polytopes of the Bravais lattice types are characterized by required zeros
of scalars and/or multiplets of equal values. For instance, all primitive 
monoclinic cells have two 90 degree angles and so two zero scalars. Consider R1 as
an example for multiplets. The character for R1 is $ \bf{rrr sss} $; two triplets.

For the case of M4, primitive monoclinic, the character is $ \bf{00r stu} $. 
For a particular value for r,s,t,u, the best monoclinic cell will not
depend on whatever values are substituted for the zeros. In the plane
corresponding to the two zero scalars, we can draw a circle, each point
on the circle corresponding to a lattice. Reducing each lattice and calculating
the distance of the reduced cell from $ \bf{00r stu} $ will give a curve that
should not have any discontinuities. See Figure \ref{fig:M4}.

Similarly, Bravais type O5 has a triplet of zeros, character $ \bf{000 rst} $. Here we
can generate a sphere that surrounds the orthorhombic point. See Figure \ref{fig:O5}.

Bravais type O4 has two zeros and a pair of required equal values, 
character $ \bf{00r sst} $. The zeros can be treated the same as above. 
The pair of equal values must 
have an average value that remains equal to $\bf{s}$ in the lattice character. 
That leaves 3 values to vary, and we can create a sphere around 
the point $ \bf{00r sst} $. See Figure \ref{fig:O4}.

\section{Comparisons}

Comparison to other distance measure; Data for other measures taken
from \citeasnoun{andrews2014}.

>>>>

\newcommand{\SVI}[0]{$\mathbf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\scalarsub}[2]{$#1_#2$}
\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\charseq}[1]{({\bf #1})}

@@

	{\LARGE \emph{\today}} \\
	\title{Determining Bravais Lattice Types}
	%\shorttitle{SELLA  Determining Bravais Lattice Types}

@@

	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
	\author[b]{Herbert J.}{Bernstein}
	\author[c]{Nicholas K.}{Sauter}

@@

	\aff[a]{9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}, Rochester Institute of Technology, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
	\aff[c]{Lawrence Berkeley National Laboratory, 1 Cyclotron Rd., Berkeley, CA, 94720 \country{USA}}

@@

	\begin{abstract}
		We introduce a new Bravais lattice determination algorithm. SELLA is a 
		straight-forward algorithm and a program for determining Bravais lattice type 
		based on Selling (Delone) reduction. It provides a clear metric of fit to each type. The mathematical foundations for treating near-symmetries are
		described. The 24 Bravais lattice types described by \citeasnoun{Delaunay1932} 
		are described as linear manifolds in 6-dimensions.
	\end{abstract}

@@

	We introduce a new Bravais lattice determination algorithm. The Bravais 
	lattice types were created by \citeasnoun{Bravais1850}. \citeasnoun{Delaunay1932} 
	and \citeasnoun{Niggli1928} developed methods for the identification 
	of the Bravais lattice type of a crystal using the measured unit cell 
	dimensions; their methods were exact only if the cell parameters 
	exactly corresponded to the actual type \cite{Patterson1957}. 
	Delaunay \cite{Delaunay1932} discussed the issues of scalars that have nearly zero values, 
	but the broader issues of measurement errors were not 
	discussed. (Delaunay later changed the spelling to Delone, a
	transliteration of the Cyrillic spelling of his name.)

	\citeasnoun{Zimmermann1985} described the program, DELOS
	for the same purpose and based on the use of Selling scalars, which
	SELLA uses as well; they treat the scalars as a list of 6 numbers, but do not consider them as points in 
	a space.  SELLA treats the scalars as elements of vectors in the 6-dimensional
	space, \SVI. Treating lattices as points in a liner space accesses the tools
	and systematic procedures of linear algebra.

	\citeasnoun{andrews2014} reviewed the literature on 
	 efforts to create methods to utilize data that contain unavoidable 
	measurement error and adds a more generic view of the problem.

	SELLA is a based on Selling (Delone) reduction. It is a complete, closed solution, and it provides a clear metric of fit to each type.

@@

	\subsection{The Bravais types}

	\citeasnoun{Delaunay1932} found that the 14 Bravais lattice types 
	could be allocated among 24 types that are determined by the 
	type of Dirichlet unit cell (termed the Voronoi region \cite{Voronoi1908} or 
	Dirichlet region \cite{dirichlet1850} by 
	Delone). Delone used the convention of \citeasnoun{Bravais1850}
	using the sides of the tetrahedron labeled by the symbols of the six
	Selling scalars. Figure \ref{fig:PQRSTU} shows the labeling used in
	the International Tables for Crystallography \cite{Henry1952}; 
	Table \ref{table:ScalarConcordance} shows other choices that have been used. The
	contents for each type are described in Figure \ref{fig:PQRSTU}.

	\citeasnoun{Burzlaff1985} revised the labeling of the types.
	Here we have returned to the numbering of the Bravais types to that 
	of \citeasnoun{Delaunay1932}.
	In some cases, Delone included two forms within a single 
	type. \citeasnoun{Burzlaff1985}
	renumbered by ordinals. We have labeled the Delone 
	types that have multiple items as "A" and "B". Table \ref{table:DeloneTypeConcordance}
	shows the various labelings. Figure \ref{fig:rotatedDeloneTypes} 
	revises the figure of \citeasnoun{Delaunay1932}, using 
	more modern notation. Table \ref{table:ScalarConcordance} 
	lists the various labelings that have been 
	used for the sides of the Bravais tetrahedron. 
	Figure \ref{fig:PQRSTU} shows the arrangement of the 
	symbols as used in the International Tables for Crystallography.

@@

			\rotatebox{80}{\citeasnoun{Delaunay1932}}
			& \rotatebox{80}{\citeasnoun{Henry1952}  }
			& \rotatebox{80}{\citeasnoun{Patterson1957} }
			& \rotatebox{80}{\citeasnoun{Burzlaff1985}}
			&\rotatebox{80}{\citeasnoun{andrews2019} } \\	
			\midrule
			g&P&h$_{23}$&$a_2.s_3$&\scalarsub{s}{1}\\		
			k&R&h$_{12}$&$a_1.a_3$&\scalarsub{s}{3}\\		
			l&S&h$_{14}$&$a_1.a_2$&\scalarsub{s}{4}\\		
			m&T&h$_{24}$&$a_1.a_4$&\scalarsub{s}{5}\\		
			h&Q&h$_{13}$&$a_2.a_4$&\scalarsub{s}{2}\\		
			n&U&h$_{34}$&$a_3.a_4$&\scalarsub{s}{6}\\		
			\bottomrule
		\end{tabular}

@@

	\begin{table}
		\begin{tabular}{ccccccccc}
			\toprule
			\cite{Delaunay1932} & \cite{Burzlaff1985}   &this  \\
			&  & paper \\
			\midrule
			K$_I$ & K1 & C1 & cI \\		
			K$_{III}$ & K2 & C3 &cF \\		
			K$_V$ & K3 & C5 &cP \\		
			Q$_I$ & Q1 & T1 &tI \\		
			Q$_{II}$ & Q2 & T2 &tI \\		
			Q$_V$ & Q3 & T5 & tP\\		
			R$_I$ & R1 & H1 & hR \\		
			R$_{III}$ & R2 & H3 & hR \\		
			O$_I$ & O1 & O1A & oF  \\		
			O$_I$ & O2 & O1B &oI  \\		
			O$_{II}$ & O3 & O2 & oI   \\		
			O$_{III}$ & O4 & O3$^*$ & oI   \\		
			O$_{IV}$ & O5 & O4 & oS   \\		
			O$_V$ & O6 & O5 & oP   \\		
			M$_I$ & M1 & M1A & mC  \\		
			M$_I$ & M2 & M1B & mC \\		
			M$_{II_2}$  & M3 & M2A$^*$ & mC \\		
			M$_{II}$ & M4 & M2B$^*$ & mC  \\		
			M$_{III}$ & M5 & M3 & mC  \\		
			M$_{IV}$ & M6 & M4 & mP  \\		
			T$_I$ & T1 & A1 & aP  \\ 		
			T$_{II}$ & T2 & A2$^*$ & aP \\		
			T$_{III}$ & T3 & A3$^*$ & aP \\		
			H$_{IV}$  & H1 & H4 & hP \\ 
			\bottomrule
		\end{tabular}
		~\\ ~\\
		\caption{The 24 Delone types. \cite{Delaunay1932}\\
			$^*$ not a full Bravais type.}
		\label{table:DeloneTypeConcordance}
	\end{table}

	\subsection{The space \SVI{}}
	\cite{andrews2019} recast the Selling parameters to define a metric space, \SVI{}.
	The base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell
	are used to define $\bf{d}=-(\bf{a}+\bf{b}+\bf{c})$.
	A point ${s}$ in \SVI{} is defined as:\\
	${s}$= [\scalarsub{s}{1}, \scalarsub{s}{2}, \scalarsub{s}{3},\scalarsub{s}{4}, \scalarsub{s}{5}, \scalarsub{s}{6}] \\
	where \scalarsub{s}{1}=\vdotv{b}{c}, ~\scalarsub{s}{2}=\vdotv{a}{c}, ~\scalarsub{s}{3}=\vdotv{a}{b}, 
	~\scalarsub{s}{4}=\vdotv{a}{d}, ~\scalarsub{s}{5}= \vdotv{b}{d}, and \scalarsub{s}{6}=\vdotv{c}{d}.

	\subsection{Terminology of the Bravais lattice types}
	The Bravais tetrahedron symbol is described in Figure \ref{fig:PQRSTU}.\\
	The different ways that the scalars have been labeled is shown in Table \ref{table:ScalarConcordance} \\
	The table of the Delone types is displayed in Figure \ref{fig:rotatedDeloneTypes}. The table has been transposed from that of Delone. Where Delone listed 2 types in one cell, we have been split them into parts A and B. The crystal family names and abbreviations have been changed to modern usage. \\

	\subsection{Character}
	In \SVI{} for reduced cells, each Bravais type is a single, connected,
	origin-inclusive, Euclidean manifold. As such, all points in a
	particular manifold can be described by their ``character'', 
	a representation of the vectorsns of that type. The table of \citeasnoun{Delaunay1932} included this information 
	encoded in the formalism of Bravais, as a tetrahedron, with
	the sides marked to show which of the 6 Selling parameters were
	zero or which pairs (or multiplets) have equal values.

	As an example, in Figure \ref{fig:O4_terminology}, the tetrahedron shows two
	zeros, in positions P and Q (\scalarsub{s}{1} and \scalarsub{s}{2}). The slashes on the lines
	show that positions S and T (\scalarsub{s}{4} and \scalarsub{s}{5}) are equal. This is 
	encoded in the \textit{character}, which is shown at the bottom of the item as \charseq{00r sst}.

	\section{Algorithms for creation of an exhaustive list of the manifolds of the Bravais types}

@@

	We will need to create projectors for each of the Bravais
	lattice types. The projectors will be use to generate
	images of probe points in each of the types, and the perps
	will measure the distance from each manifold.

	We will use Selling reduction; Niggli reduction has a fundamental 
	unit that is non-convex and some boundaries of the fundamental 
	unit are partly closed and partly open \cite{andrews2014} leading 
	to considerable complexity. Because Niggli reduction divides the 
	fundamental unit of Niggli reduction into separate all obtuse and all acute regions, 
	the transitions between those are quite complex \cite{andrews2014}. 
	In space \SVI{} \cite{andrews2019b}, the fundamental 
	unit is the simple all-negative orthant of a 6-dimensional 
	Cartesian space. Only two kinds of operations in the space 
	will need to be considered: the 6  transform operations 
	at the boundaries of the orthant (zeros of an axis) and 
	24 reflections (allowed permutations) \cite{andrews2019}

	Twenty-four lattice types were enumerated by \citeasnoun{Delaunay1932}. 
	Three of those are triclinic, and we will ignore them because all crystals 
	can be indexed as triclinic. We 
	shall need all representations of all non-triclinic lattice 
	types within the \SVI{} fundamental unit (all Selling-reduced 
	cells) or those generated beyond the boundaries by boundary transforms. 
	Three forms (M3, M2B, and O3) are reduced dimension boundaries that do
	not correspond to unique Bravais types; any lattice that falls in those
	three is indexed by one of the other types.

	Seven of the 21 Delone types do not have a zero in their \SVI{} 
	representation; the presence of zero indicates that a form is 
	on the boundary of the \SVI{} fundamental unit. Since a 
	point on the boundary transforms to another boundary 
	point \cite{andrews2019b}, we can generate the Virtual 
	Cartesian Point (VCP) \cite{andrews2019b} appropriate to 
	each zero in the \SVI{} vector. Because the reduction
	operations do not commute, in the cases where there is more 
	than one zero in the vector, all orderings of applying 
	reductions to all of the zeros must be generated. (For example, 
	for two zeros, there will be six vectors produced.) Again, 
	there may be many duplicate results.

	Finally, the projectors must be computed for each of the many 
	sample vectors of the 21 types. We will also need the 
	``perps'', the operators that give the normal vector to each 
	such manifold. For a particular manifold, the projector 
	applied to the probe gives
	the least-squares best fit point within that manifold
	to the probe vector. 
	Perps are computed by subtracting the 
	corresponding projector from the unit matrix. The norm of avector
	created by multiplying the probe times the projector
	 is the distance from that manifold.

	\subsubsection{Generating sample vectors from all Bravais types\protect\\}

	Step 1: Generate any random vector that contains no zero 
	values and no duplicate values.  For a later operation, 
	it should only contain values large enough that adding or
	subtracting the smallest
	representable floating point number will not change the 
	value in computer arithmetic ($DBL\_MIN$ in the C language).
	That value will be used to mark the positions of zeros
	in vectors that have been processed.

	Step 2; For each Delone type, multiply its projector times 
	the vector from Step 1. Store  the resulting vector
	from Step 1 in a list 
	with entries for each Bravais type (see Table \ref{table:DeloneTypes} 
	for the projectors).

@@

	a) If the vector result of Step 2 contains no zeros, generate the 24 reflections
	and store in the list by Bravais type.

	\textit{\textbf{LCA--need to normalize the vectors, just like 
		in lattice matching. ? Can we use the 
		lattice matching code here? This is very 
		incomplete. The processing of multiple zeros is MISSING! But the use of DBL\_MIN is there.}}

	b) If the vector contains one zero, apply the boundary transform 
	operation for that boundary
	to the vector, and
	store the products of the vector and its 24 reflections in the list. Note, 
	this is where the choice of $DBL\_MIN$ has effect; adding or
	subtracting such a small value will not change
	any value except zero.

	c) If the vector contains more than one zero and \underline{not} $DBL\_MIN$, 
	first set one zero to $DBL\_MIN$,  and then return it to 
	step b, using the modified vector. Then set the second zero to $DBL\_MIN$ (with the
	first zero still zero), and send that result to step b. 
	If there is a third zero,  process it the same way as 2 zeros and return to c.
	(Note that no valid unit cell can have more than 3 zeros
	in space \SVI{}.)

@@

	\subsubsection{Generate projectors from sample vectors\\}

	Each vector in each of the lists from the previous steps is a representative of one
	of the Delone types. Although a vector is a one-dimensional manifold,
	these vectors encode the information about the full manifolds of the
	Delone types in their zeros and their duplicate values. Now we 
	discuss the creation of the projector matrices that correspond
	to each vector.

@@

	\textit{\textbf{LCA write pseudocode for Fraction.}}

	The following steps will be repeated for each of the vectors in the
	list generated above in order to generate the list
	of the corresponding projector matrices.

	To generate a matrix \textit{m} starting from a vector.

	Step 1: Zero any elements of the vector that have been set to $DBL\_MIN$.

@@

	\subsection {Measuring the fit}

	When the library of projectors has been prepared, then
	the distance of any probe from each of the manifolds
	can be computed. 
	The distance from the given manifold is just
	the norm of the perp for that manifold applied to the probe.
	An alternative, scale-free measure is the angle between
	the sample vector and its projected image.

@@

	Three of the 21 non-triclinic Delone types are only boundaries of other types. In Figure 
	\ref{fig:degenerate}, Delone types M2B and M3 are in the row for orthorhombic, and
	O3 is in the row for rhombohedral and tetragonal. These types do not need to
	be searched for in general surveys. If a probe is close to 
	one of these degenerate manifold, it will be close to one
	or both of the adjacent manifolds.

@@

	Most of the proposals for Bravais lattice determination in the 
	presence of experimental error do not have a clear proof that 
	they are complete. For many years there were verbal complaints 
	that available programs would not infrequently miss experimental 
	cases (for example, TRACER \cite{lawton1965} and many anecdotal 
	communications).  To our knowledge, only the method of 
	\cite{OishiTomiyasu2012} has clearly demonstrated completeness
	for exact values.

	The justification that Sella is complete has three parts.

	First: We require that the cell being tested be Selling reduced, and
	all Bravais lattice types within the \SVI{} fundamental unit (all
	\SVI{} scalars negative) to have their projectors generated. Therefore,
	the 24 possible forms (reflections) of a 
	reduced cell will be close to one of the 24 copies of the types. 
	For the reflections, we do not need to consider the case of 
	points near a boundary; the boundary 
	transforms must be treated for that.

	Second: In the case of the boundary transform operations, 
	we take the case of a point that is infinitesimally distant from the manifold of a 
	Bravais lattice type that has one zero in its scalars. Take 
	the point in the manifold that is closest to the experimental 
	point. We transform both by the corresponding boundary transform 
	operation. We have generated a new pair of points in a 
	different location, one of which is in some representation of 
	the same Bravais lattice type. The points are still 
	infinitesimally distant from each other.

	Third: Assume we have a point that is outside the fundamental 
	unit and is near an undiscovered manifold for a Bravais lattice 
	type. If the point is near a boundary defined by a single 
	zero, then we apply a boundary transform to the point and to 
	the Bravais manifold, bringing them both into the fundamental 
	unit. If all of the variations of all of the Bravais 
	lattice types within and on the boundaries of the fundamental unit have 
	been generated, then this putative undiscovered manifold 
	has already been included in the list of possibilities. 
	Similar arguments apply to the cases of two or three zeros.

	If we apply all of the boundary transform operations to the 
	representations of the Bravais lattice types, then 
	there are no places that a point in the fundamental 
	unit will fail to find the closest Bravais lattice 
	(or lattices). As stated above, we found 239 
	non-triclinic possible representations.

	What might occur if some reduced point has an unreduced 
	copy near the all-negative orthant?  Would it be 
	possible for it to be close to one of the copies of one 
	of the 21 Delone types? The reduced point is obviously in 
	the non-positive orthant. Further, because all possible 
	copies of all of the Delone type manifolds have been 
	generated, the reduced point will be near a copy of the 
	manifold it was near when it was unreduced.

@@

	89539 unit cells were extracted from the Protein Data Bank 
	in 2018
	(\citeasnoun{bernstein1977} and \citeasnoun{berman2000}). 
	When evaluated with Sella, all but 57 were found at zero 
	distance from the described crystal class. The 57 were 
	identified as incorrect unit cells for the assigned 
	crystal class; however, the distances from the assigned 
	classes were small because the deviations from the 
	appropriate parameters were modest (see Table \ref{tab:PDBCells}).
	The suspect cases have since been corrected \cite{Burley_personal_communication}.
	\begin{table}

@@

		\caption{57 unit cells from the PDB around 2018 that did not match their types (later corrected)}
		\label{tab:PDBCells}
	\end{table}
	\subsection{Testing for continuity}

	The manifolds of the Bravais lattice types are characterized by required zeros
	of scalars and/or multiplets of equal values. For instance, all primitive 
	monoclinic cells have two 90 degree angles and so require two zero scalars. Consider H1 as
	an example for multiplets. The character for H1 is \charseq{rrr sss}; two triplets of equal values.

	Consider the case of M4, primitive monoclinic, the character is \charseq{00r stu}. 
	For particular values for r,s,t,u, the best monoclinic cell will not
	depend on whatever values are substituted for the zeros. (It does not depend on the zero values because the projectors to that 
	manifold create zero elements for any vector.) Imangine a plane
	with axis that are the values of the two elements of the character that are zero. Points in the plane represent unit 
	cells with angles defined by the \scalarsub{s}{1} and \scalarsub{s}{2} and r,s,t,u. We can draw a circle in the
	\scalarsub{s}{1},\scalarsub{s}{2} plane, each point
	on the circle corresponding to a lattice.

	Reducing the lattices corresponding to the points on the
	circle and r,s,t,u and calculating
	the distance of the reduced cell from \charseq{00r stu}.
	Plotting the curve of the circle with the distance of
	each point from the origin changed to distance above
	 will give a curve that
	should not have any discontinuities. See Figure \ref{fig:M4}.

	Similarly, Bravais type O5 has a triplet of zeros, character \charseq{000 rst}. Here we
	can generate a sphere that surrounds the orthorhombic point. See Figure \ref{fig:O5}.
